---
title: "How does a bike-share navigate speedy success?"
author: "Eric L Sammons"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Hypothesis / Question
Based on bike share rider data, what steps could we take to convert more casual riders to a paid membership?

## Setup

```{r Install required libraries, echo=FALSE}
# install.packages("readr")
# install.packages("data.table")
# install.packages("tidyverse")
# install.packages("dplyr")
# install.packages("scales")
# install.packages("viridis")
```

### Load required libraries

```{r Load required libraries, message=FALSE, warning=FALSE}
library(readr)
library(data.table)
library(tidyverse)
library(dplyr)
library(scales)
library(viridis)
```

### Set variables

```{r Set variables, echo=TRUE}
base_url <- "https://divvy-tripdata.s3.amazonaws.com/"  # Replace with the actual URL
destination_folder <- "data/"
months <- c("202205", "202206", "202207", "202208", "202209", "202210",
            "202211", "202212", "202301", "202302", "202303", "202304", "202305")
```

## Functions required to download and validate data files

```{r Download and extract function, include=TRUE, echo=TRUE}
download_and_extract_csv <- function(url, destination_folder) {
  # Extract the filename from the URL
  filename <- basename(url)
  
  # Download the zip file
  if (!file.exists(paste0(destination_folder, filename))) {
    download.file(url, destfile = paste0(destination_folder, filename))
    
    # Extract the CSV file from the zip
    unzip(zipfile = paste0(destination_folder, filename),
          files = gsub("\\..*", ".csv", filename),
          exdir = destination_folder)
  }
}
```


```{r Check Column names funcion, include=TRUE, echo=TRUE}
check_column_names <- function(file_paths) {
  # Read the column names from the first file
  column_names <- fread(file_paths[1])$names
  
  # Check column names in all files
  all_same <- all(sapply(file_paths, function(path) {
    identical(column_names, fread(path)$names)
  }))
  
  return(all_same)
}
```

## Download and Extract data files 
In this step we will leverage the list of YYYYMM to download 12 months of
data. Files will only be downloaded if they do not already exist in the 
destination folder (/data/, by default).

```{r Download and extract data, echo=TRUE}
for (month in months) {
  url <- paste0(base_url, month, "-divvy-tripdata.zip")
  download_and_extract_csv(url, destination_folder)
}
```

## Validate the datafiles and create a single dataframe
Column names must be the same across all CSV files. If column names are not the 
same additional work will be required and is outside the scope of this effort.

```{r Check columns and build dataframe, echo=TRUE, message=FALSE, warning=FALSE}
csv_files <- list.files(destination_folder, pattern = "\\.csv$", full.names = TRUE)

if (!check_column_names(csv_files)) {
  stop("Not all files have the same column names.")
}

df <- map_dfr(csv_files, read_csv)
```

## Housekeeping

### Remove columns not required
For our analysis we will not require the following colunns:

- start_lat
- start_lng
- end_lat
- end_lng

```{r Remove columns, include=TRUE, echo=TRUE}
df <- df %>%
  select(-c(start_lat, start_lng, end_lat, end_lng))

colnames(df)
```
## Rider type summary

```{r Verify values in member_casual, include=TRUE, echo=TRUE}
table(df$member_casual)
```

## Data transformations

### Break down the started_at column

```{r Break down the started_at column, include=TRUE, echo=TRUE}
df$date <- as.Date(df$started_at)
df$month <- format(as.Date(df$date), "%m")
df$day <- format(as.Date(df$date), "%d")
df$year <- format(as.Date(df$date), "%Y")
df$day_of_week <- format(as.Date(df$date), "%A")
```
 
### Create ride_length column

```{r ride_length, include=TRUE, echo=TRUE}
df$ride_length <- difftime(df$ended_at, df$started_at)
df$ride_length <- as.numeric(df$ride_length)

str(df$ride_length)
```
### Remove rows with ride_length < 0

```{r Remove invalid data, include=TRUE, echo=TRUE}
df <- df[!(df$ride_length < 0),]
```

## Analysis

### Ride length summary

```{r Analysis, include=TRUE, echo=TRUE}
summary(df$ride_length)
```

### Ride length summary member vs casual

```{r member vs casual, include=TRUE, echo=TRUE}
comparison <- aggregate(df$ride_length, by = list(df$member_casual), FUN = function(x)
  c(mean = mean(x),
    median = median(x),
    max = max(x),
    min = min(x)))

comparison
```
### Average ride time by day of week and member type
Here we look at the mean *ride_length* grouped by *day_of_week* and *member_casual*

```{r message=FALSE, warning=FALSE}
df %>%
  mutate(day_of_week = factor(day_of_week, levels = c("Sunday", "Monday", 
                                                      "Tuesday", "Wednesday", 
                                                      "Thursday", "Friday", 
                                                      "Saturday"))) %>%
  group_by(day_of_week,member_casual) %>%
  summarise(mean = mean(ride_length))
```

### Create Analysis Dataframe
To make the next steps a bit easier we will create a new dataframe. This dataframe, rider_analysis, will focus on:

- Number of rides
- Average duration of ride (in minutes)
- By day of week
- By member type (member_casual)

```{r message=FALSE}
rider_analysis <- df %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(),
             average_ride_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)
```

### Quick Look At Average Duration vs. Number of Rides
In this plot we take a quick look at the number of rides and their average duration (in minutes) by day of the week.

```{r}
color_theme <- c("#F05039", "#7CA1CC")

ggplot(rider_analysis) +
  geom_point(mapping = aes(x = average_ride_duration, 
                           y = number_of_rides, 
                           color = member_casual)) +
  facet_wrap(~ weekday) +
  labs(x = "Average duration (minutes)", 
       y = "Number of rides", color = "Day of Week") +
  scale_color_manual(values = color_theme) +
  scale_y_continuous(labels = scales::comma)
```

### Bike Share: Number of Rides By Weekday and Rider Type
Here we visualize the number of rides by day of week and whether the rider is a member of casual.

```{r Visualize ridership}
ggplot(rider_analysis) +
  geom_col(mapping = aes(x=weekday, 
                         y=number_of_rides, 
                         fill=member_casual), 
           position = "dodge") +
  labs(title = "Bike Share: Number of Rides",
       subtitle = "By Weekday",
       y = "Number of rides", 
       x = "Day of week",
       fill = "Casual or Member") +
  scale_x_discrete(limits = rev(c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"))) +
  scale_y_continuous(labels = scales::comma, 
                     limits = c(0,max(rider_analysis$number_of_rides) + 100000)) +
  theme(axis.text.x = element_text(angle = 25)) +
  coord_flip()
```

### Bike Share: Average Duration By Weekday
Here we visualize the average duration of rides by day of week and whether the rider is a member of casual.

```{r}
ggplot(rider_analysis) +
  geom_col(mapping = aes(x = weekday, 
                         y = average_ride_duration, 
                         fill = member_casual),
           position = "dodge") +
  labs(title = "Bike Share: Average Duration",
       subtitle = "By Weekday",
       y = "Average ride duration (minutes)",
       x = "Day of week", 
       fill = "Casual or Member") +
  scale_x_discrete(limits = rev(c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"))) +
  scale_y_continuous(labels = scales::comma, 
                     limits = c(0,max(rider_analysis$average_ride_duration))) +
  theme(axis.text.x = element_text(angle = 25)) +
  coord_flip()
```

## Conclusion
In conclusion we find the following that casual riders are riding longer, on average, vs their member counterparts. However, we see that there are far more member riders than casual riders on most days of the week. On weekends we do see that the number of casual rides vs member rides are nearly the same.

## Recommendation
Based on our findings here the following recommendaions may help convert casual riders to a paid membership.

- Introduce a weekend only membership that includes discounts for riding on weekdays.
- Introduce a discount program for members who ride longer.